---
title: 'Project: `r params$project`'
subtitle: 'Screen: `r params$screen`'
author: 'Type: `r params$hts_type`'
date: 'Date: `r format(Sys.Date(), "%d-%m-%Y")`'
format:
  html:
    fig-width: 8.0
    fig-height: 6.0
  pdf:
    fig-width: 7.0
    fig-height: 5.0
  docx:
    fig-width: 7.0
    fig-height: 5.0
    reference-doc: mystyles.docx
editor_options:
  chunk_output_type: console
params:
  project: 2023_snl_sartori_biochemical
  screen: 250122_screen_report
  hts_type: single
  condi_yes: MC38
  condi_no: SW620
  geom_cor: 'no'
  act_cut: log10(1.5)
  fdr_cut: 0.01
  select_cut: 0.3
  select_cut_yes: 2 * act_cut
  select_cut_no: act_cut
  path_data: C:/Users/mprummer/Downloads/250113 Screen NEXUS LLD-S_Label2 (1).csv
  path_meta: C:/Users/mprummer/Downloads/experiment_data (1).csv
  path_lib: C:/Users/mprummer/Downloads/250113 Screen NEXUS LLD-S_Label2.csv
  path_output: C:/Users/mprummer/Downloads/
  path_SLmisc: C:/Users/mprummer/OneDrive - ETH Zurich/00_NEXUS/Statistics/Projects/2025_internal_LDM/
execute:
  echo: false
  warning: false
  message: false
---



```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo = F, error=F, message=F, warning=F, cache=F, results='hide', fig.show = "hide")
```


```{r}

# params = list(
#   project = "2024_snl_suter_PARP7",
#   screen = "20250513 Screen Batch 1",
#   hts_type = "single", # c("single", "selectivity", "doseresponse")
#   geom_cor = "median polish", # c("no", "median polish", "local polynomial fit", "loess")
#   act_cut = "log10(1.5)", # must be a string
#   fdr_cut = 0.01, # [0, 1]
#   select_cut = 0.3, # [0, oo]
#   select_cut_x = "2 * act_cut", # target act, must be a string
#   select_cut_y = "act_cut", # counter act, must be a string
#   # path to measurement data "MAIN INFO"
#   path_data = "C:/Users/mprummer/OneDrive - ETH Zurich/00_NEXUS/Statistics/Projects/2024_snl_suter_PARP7/20250513 Screen Batch 1/20250513 Screen Batch 1_Lum_CTG_chemical_info.csv",
#   # path to experiment description "EXPERIMENT DATA"
#   path_meta = "C:/Users/mprummer/OneDrive - ETH Zurich/00_NEXUS/Statistics/Projects/2024_snl_suter_PARP7/20250513 Screen Batch 1/experiment_data-4.csv",
#   # path to library description "CHEMICAL INFO"
#   path_lib = "C:/Users/mprummer/OneDrive - ETH Zurich/00_NEXUS/Statistics/Projects/2025_internal_LDM/241126_Screen_NexusApproved_SW620_Lum.csv",
#   # output path for all result files
#   path_output = output_dir
# )

```


```{r}
#| label: init

#####################################################################################
## NO USER INPUT REQUIRED
#####################################################################################
source(paste0(params$path_SLmisc, "SLmisc.R") )

# packages
lby = c("RColorBrewer", "ggplot2", "gplots", "ggrepel", "cowplot", 
        "ggpubr", "ggExtra", "dplyr", "reshape2", "lubridate",
        "reshape2", "knitr", "pheatmap", "tidyr", "locfit", "robust","MASS")
# no lib.loc: in the Docker image MASS is in the system library, not in .libPaths()[1]
resp = lapply(lby, require, character.only=T, warn.conflicts=F, quietly=T)
if(!all(unlist(resp))) stop("Could not load one or more required packages")
rm(resp)

# functions
'%&%' = function(a,b) paste(a,b,sep="")
theme_set(theme_bw(12))

# custom palettes
type.pal = brewer.pal(9,"Set1")
type.pal = c(type.pal[2:1], "black", type.pal[3:9], brewer.pal(8,"Dark2"))

# read in parameters
hts_type = params$hts_type #c("single", "selectivity", "doseresponse")[1]
lib_type = params$lib_type #c("actitarg", "nexus_fda", "selleckchem_clinical", "LLD")[4]
geom_cor = params$geom_cor #c("no", "median polish", "local polynomial fit", "loess")[1]
act_cut = eval(parse(text=params$act_cut))
fdr_cut = params$fdr_cut
select_cut = params$select_cut
path_data = params$path_data
path_meta = params$path_meta
path_libfile = params$path_lib
path_output = params$path_output

```



```{r}
#| label: load_data

d_in = read.LDM(path_data)
id_keep = which(!duplicated(d_in))
d_in = d_in[id_keep, ]
meta_plates = read.ExpData(path_meta)
id_keep = which(!duplicated(meta_plates))
meta_plates = meta_plates[id_keep, ]
dd = left_join(d_in, meta_plates)
```


```{r}
#| label: screen_specific_cleaning

if(file.exists(path_output %&% "clean.R")) {
  source(path_output %&% "clean.R", local=myEnv)
}
```



```{r}
#| label: clean_data

dd = dd[!is.na(dd$cpd_type), ]
dd$meas = dd$measurement_label
dd$cpd_type = factor(dd$cpd_type, levels = union(c("N", "P", "C"), unique(dd$cpd_type)))
dd$meas = factor(dd$meas)
dd$lib_plate = factor(dd$lib_plate, levels=unique(dd$lib_plate))
dd$replicate = factor(dd$replicate, levels=unique(dd$replicate))
dd$cell_type = factor(dd$cell_type, levels=unique(dd$cell_type))
dd$condition = factor(dd$condition, levels=unique(dd$condition))
dd$layout = factor(dd$layout, levels=unique(dd$layout))
```


```{r}
#| label: plate_order

dd$meas_ts = as_datetime(dd$measurement_timestamp)
dd$plate = factor(dd$plate, levels=unique(dd$plate[order(dd$meas_ts)]))
```


# Objectives 

...

# Assay description

...

### Test conditions & compound libraries

...


# Experimental design


We obtained a data description file with the layout for each plate for which a summary is shown in @tab-meta_descr. 

```{r}
#| label: tbl-meta_descr
#| results: asis
#| tbl-cap: "Description of the meta data table. The content of this table shows the kind and amout of data expected from the design."

kable(colNames.colValues(meta_plates, 80), 
      rownames=F)
```


# Analysis plan 

### Preprocessing

Systematic variation from plate to plate is removed by standard normalization procedures[^Malo]. Following common practice, assay quality is evaluated on the basis of the SSMD and the Z' factor. Within-plate row-wise or column-wise stripe patterns as well as edge effects can arise during plate preparation. These patterns are eliminated using the median polish method of Tuckey[^Tukey] or by subtracting a smooth polynomial using the loess function[^Boutros].

[^Malo]: Malo N, Hanley JA, Cerquozzi S, Pelletier J, Nadon R (2006) "Statistical practice in high-throughput screening data analysis" _Nat Biotech_ **24**, 167-175. DOI: 10.1038/nbt1186.
[^Tukey]: Tukey, J. W. (1977). "Exploratory Data Analysis", Reading Massachusetts: Addison-Wesley.
[^Boutros]: Boutros M, Brás LP, Huber W (2006). "Analysis of cell-based RNAi screens" _Genome Biology_ **7**, R66. DOI: 10.1186/gb-2006-7-7-r66.


### Differential activity analysis 

Differential activity was analyzed following the workflow sketched in [^prummer]. Briefly, for each single-dose measurement of a compound, a Z-test is performed against the Null hypothesis that its activity is indistinguishable from the negative controls. For this to be valid, the distribution of activities of the negative controls are checked to be normal. The mean and variance of the distribution is estimated robustly for each plate and, if appropriate, smoothly averaged over a range of consecutive plates. 

The semi-automated workflow is implemented in the _R environment for statistical computing_[^biocref].

[^prummer]: Prummer M (2012) "Hypothesis Testing in High-Throughput Screening for drug discovery" _J Biomol Screen_ **17**, 519-529. DOI: 10.1177/1087057111431278.
 
[^biocref]: Huber W et al. (2015) "Orchestrating high-throughput genomic analysis with Bioconductor" _Nat Methods_ **12**, 115-121. DOI: 10.1038/nmeth.3252.




# Data description

```{r}
#| label: compute_stats

assay_plates = unique(dd$plate)
nr_assay_plates = length(assay_plates)
lib_plates = unique(dd$lib_plate)
nr_lib_plates = length(lib_plates)
layouts = unique(dd$layout)
nr_layouts = length(layouts)
experiments = unique(dd$experiment)
nr_experiments = length(experiments)
replicates = unique(dd$replicate)
nr_replicates = length(replicates)
conditions = unique(dd$condition)
nr_conditions = length(conditions)
measurements = unique(dd$measurement_label)
nr_measurements = length(measurements)
cell_types = unique(dd$cell_type)
nr_cell_types = length(cell_types)
```


We received data files from `r nr_assay_plates` plate(s), with `r nr_measurements` measurement(s) per well: 


```{r}
#| label: tbl-plate_table
#| results: asis
#| tbl-cap: "List of plates in the screen and their size."

tmp = table(dd$plate, dd$meas)

if(length(dim(tmp)) == 1){
  tmp = data.frame(plate=names(tmp), size=as.numeric(tmp), stringsAsFactors = F)
} else {
  tmp = melt(tmp)
  names(tmp) = c("plate", "measurement", "size")
}
tab = tapply(as.character(tmp$plate), tmp$size, concat.chr.max, nchar_max=80)
tab = data.frame(Barcode=as.character(tab), Size=names(tab))
kable(tab)
```


```{r}
#| label: tbl-screen_summary
#| results: asis
#| tbl-cap: "Number of wells for the different compound types."

tab = as.matrix(table(dd$cpd_type))
colnames(tab) = "Nr of wells"
# tab = dd[dd$channel == levels(dd$channel)[1], ]
# tab = cbind(as.matrix(table(tab$type, tab$plate)), "Total" = as.numeric(table(tab$type)))
kable(tab)
```

## Plate layout


```{r}
#| label: fig-plate_layout
#| fig-show: asis
#| fig-width: 6
#| fig-height: 3.5
#| fig-cap: "Heatmap representation of compound well allocation. "
#| fig-dpi: 300

col.pal = colorRampPalette( rev(brewer.pal(9, "Blues")) )(length(unique(dd$conc)))
col.pal = c(type.pal[seq_along(unique(dd$cpd_type))], col.pal)
p1 = ggplate.hm(dd, "cpd_type", plot=F)
plot(p1  + facet_wrap(~layout, scales="free", nrow = 2))
```


# Preprocessing, QC

## Control wells


```{r}
#| label: fig-controls_nonorm
#| fig-show: asis
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Raw signal of control wells."
#| fig-dpi: 300

nr_rows = floor(sqrt(length(levels(dd$meas))))
ggplot(dd[dd$cpd_type!="C", ], aes(x=plate, y=raw, group=cpd_type, color=cpd_type)) +
  geom_jitter(width=rel(0.2), show.legend = T) + ylab("Raw signal") + xlab("") +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas, nrow=nr_rows, scales="free_y") +
  guides(color=guide_legend(ncol=1, byrow=T)) +
  #coord_cartesian(ylim=c(0,1.1*max(tmp$value, na.rm=T))) + 
  theme(strip.text = element_text(size=10), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size=rel(0.99)),
        legend.position="right",
        legend.text=element_text(size=rel(0.8)))

```

## Transformation

The raw signal is of heteroscedastic nature, i.e., the variance is not uniform but dependent on the signal strength. This makes interpretation of signal to noise levels difficult. We therefore transform the raw signal according to

$$ y = log_{10}(raw) $$


```{r}
#| label: fig-controls_trafo
#| fig-show: asis
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Transformed signal of control wells."
#| fig-dpi: 300

dd$trafo = log10(1 + dd$raw)

ggplot(dd[dd$cpd_type!="C", ], aes(x=plate, y=trafo, group=cpd_type, color=cpd_type)) +
  geom_jitter(width=rel(0.2), show.legend = T) + ylab("Transformed signal") + xlab("") +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas, nrow=1, scales="free_y") +
  guides(color=guide_legend(ncol=1, byrow=T)) +
  theme(strip.text = element_text(size=10), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size=rel(0.99)),
        legend.position="right",
        legend.text=element_text(size=rel(0.8)))

```

## Normalization 

Percent-control normalization according to

$$ Z = \frac{y - median(y_N)}{median(y_P) - median(y_N)}$$

```{r}
#| label: plate_qc

plate_stats = get_plateStats(dd, "trafo")
plate_stats$label = gsub(".*_(.*)", "\\1", plate_stats$plate)

ggplot(plate_stats, aes(x=med, y=mad, color=cpd_type, label=label)) + 
  geom_point(alpha=0, show.legend = T) + 
  geom_abline(slope=1, intercept=0, linetype=2) +
  geom_label(size=3, show.legend = F) +
  facet_wrap(~meas, scales = "free")+
  scale_color_manual(name="", values=type.pal, drop=F) +
  guides(color = guide_legend(override.aes = aes(label = "", alpha = 1))) 

tmp = melt(plate_stats, measure.vars = c("med", "mad"))
plate_stats = dcast(tmp, plate + meas + label ~ variable + cpd_type, 
                    value.var = "value")
plate_stats$z_prime = 1 - 3*(plate_stats$mad_N + plate_stats$mad_P) / 
                         abs(plate_stats$med_P - plate_stats$med_N)
plate_stats$SSMD = abs(plate_stats$med_P - plate_stats$med_N) / 
                sqrt(plate_stats$mad_N^2 + plate_stats$mad_P^2)

dd = left_join(dd, plate_stats)
dd$norm = (dd$trafo - dd$med_N) / (dd$med_P - dd$med_N)

```

```{r}
#| label: fig-ctr_norm
#| fig-show: asis
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Log-transformed signal of compound wells before (trafo) and after normalization (norm)."
#| fig-dpi: 300

tmp = melt(dd, measure.vars = c("trafo", "norm"), 
           value.name = "value", variable.name = "signal")
ggplot(tmp[tmp$cpd_type=="C", ], aes(x=plate, y=value, group=cpd_type, color=cpd_type)) +
  geom_jitter(width=0.2, show.legend = T) + ylab("Signal") + xlab("") +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_grid(signal ~ meas, scales="free_y") + 
  guides(color=guide_legend(ncol=1,byrow=T)) +
  theme(strip.text = element_text(size=10), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size=rel(0.99)),
        legend.position="right", legend.title=element_text(size=rel(0.8)),
        legend.text=element_text(size=rel(0.8)))
```


## Screen quality: Z' and SSMD

The Strictly Standardized Mean Difference is defined as

$$ SSMD = \frac{\mu_P - \mu_N}{(\sigma_P^2+\sigma_N^2)^{-1/2}},$$
and the _Z-prime factor_ as

$$ Z'= 1- \frac{6 \bar{\sigma}}{\Delta\mu},$$
with $\bar{\sigma} = 1/2\cdot(\sigma_P+\sigma_N)$ the average standard deviation and $\Delta\mu = \mu_P - \mu_N$ the difference in mean of the positive and the negative control. Both are standard QC measures to judge the quality of entire screens and make comparison between screens. 

```{r}
#| label: fig-zprime
#| fig-show: asis
#| fig-width: 7
#| fig-height: 3
#| fig-cap: "Plate-wise QC parameters. (A) Z' factor (B) SSMD. The colored background indicates the ranges for good (green), medium (yellow) and bad quality (red)."
#| fig-dpi: 300

cols=brewer.pal(9,"Set1")[c(1,6,3)]
y_lims = c(min(0, min(plate_stats$z_prime)) - 0.1, 1)
p1 = ggplot(plate_stats, aes(x=plate, y=z_prime)) + ylab("Z' factor") + xlab("") +
  #scale_y_continuous(limits=c(-0.2,1)) + 
  geom_hline(yintercept=c(0,0.5)) +
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=-Inf, ymax=0, fill=cols[1]) + 
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=0, ymax=0.5, fill=cols[2]) + 
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=0.5, ymax=1, fill=cols[3]) + 
  geom_point() + 
  scale_y_continuous(limits = y_lims) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        strip.text = element_text(size=10))
#ggMarginal(p1, type="histogram", margins="y")
p2 = ggplot(plate_stats, aes(x=plate, y=SSMD)) + xlab("") +
  #scale_y_continuous(limits=c(-0.2,1)) + 
  geom_hline(yintercept=c(0,0.5)) +
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=0, ymax=3, fill=cols[1]) + 
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=3, ymax=6, fill=cols[2]) + 
  geom_rect(xmin=0, xmax=length(levels(tmp$plate))+1, ymin=6, ymax=Inf, fill=cols[3]) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, max(plate_stats$SSMD)+0.1)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        strip.text = element_text(size=10))
plot_grid(p1, p2, nrow=1, labels = "AUTO")

```

```{r}
#| label: tbl-zprime_stats
#| results: asis
#| tbl-cap: "Comprehensive Z' statistics for the screen."

tab = plate_stats$z_prime
tab = quantile(tab, c(0,0.05,0.5,1), na.rm=T)
tab = as.matrix(tab)
tab = rbind(tab,length(which(plate_stats$z_prime>0.5)))
colnames(tab) = ""
rownames(tab) = c("Min", "P(Z' > x) = 95%", "Median", "Max", "Nr plates: Z' > 0.5")
kable(signif(tab,2))

```


```{r}

qc.string = "bad"
if(tab[2,1]> -0.1) qc.string = "borderline"
if(tab[2,1]> 0.1) qc.string = "fair"
if(tab[2,1]> 0.25) qc.string = "good"
if(tab[2,1]> 0.5) qc.string = "very good"
if(tab[2,1]> 0.75) qc.string = "excellent"

```


```{r}
#| label: plate_min_effect

plate_stats = get_plateStats(dd, "norm")
tmp = melt(plate_stats, measure.vars = c("med", "mad"))
plate_stats = dcast(tmp, plate + meas ~ variable + cpd_type, value.var = "value")

x01 = qnorm(0.99, sd = plate_stats$mad_N)
dz_min = qnorm(0.9, mean = x01, sd = plate_stats$mad_N)
ci = round(quantile(dz_min, c(0.025, 0.5, 0.975), na.rm=T), 2)
```

In summary, the overall quality of the screen was `r qc.string`. More than 95% of all plates had Z' > `r signif(tab[2,1],2)` and `r round(tab[5,1],0)` plates had $Z' > 0.5$. Based on a maximum false positive rate of 0.01 and a maximum false negative rate of 0.1, the estimated minimum detectable effect size in this screen is `r ci[2] %&% ", 95%CI [" %&% ci[1] %&% ", " %&% ci[3] %&% "]"`.



## Geometric patterns


```{r}
#| label: fig-plate_row_col
#| fig-show: asis
#| fig-width: 12
#| fig-height: 12
#| fig-cap: "Normalized signal versus column (A) or row number (B) for the first 2 plates. Systematic row or column effects can be identified if present."
#| fig-dpi: 300

plates_included = levels(dd$plate)
plates_included = plates_included[seq_len(min(2, length(plates_included)))]
tmp = dd[dd$plate %in% plates_included, ]
p1 = ggplot(tmp, aes(x=row, y=norm, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~ meas + plate, ncol=2)
p2 = ggplot(tmp, aes(x=col, y=norm, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~ meas + plate, ncol=2)

ggarrange(p1, p2, nrow=2, labels = "AUTO", common.legend = T, legend = "right")

```



```{r}
#| label: geom_cor

dd = dd[order(dd$plate, dd$row, dd$col), ]
dd$cor = dd$cor.mp = dd$cor.loc = dd$cor.loess = dd$norm

for(ii in levels(dd$plate)){
  # ii = dd$plate[1]
  td = dd[dd$plate == ii & dd$cpd_type=="C", ]
  mat = matrix(td$norm, nrow = length(unique(td$row)), byrow=F)
  m = medpolish(mat, eps = 1e-5, maxiter = 200, trace.iter=F, na.rm=T)
  dd$cor.mp[dd$plate == ii & dd$cpd_type=="C"] = as.numeric(m$residuals)
}

## assumes a column "pattern" in dd, a 2-level factor defining a pattern
## Examples: (i) stripes affecting some rows or (ii) some column, or (iii) all outer wells.
## TODO: extend to work with an arbitrary amount of regions.

if("pattern" %in% names(dd)){
  geom_cor="loess"
  dd$pattern[dd$pattern==""] = NA
  dd$pattern = factor(dd$pattern)
  region = levels(dd$pattern)
}

center_row = max(dd$row)/2
center_col = max(dd$col)/2

for(ii in levels(dd$plate)){
  # ii = dd$plate[1]
  id_select = dd$plate == ii & dd$cpd_type=="C"
  df = dd[id_select, ]
  df$xpos = df$col - center_col
  df$ypos = df$row - center_row
  if (nrow(df) > 4 & !all(is.na(df$norm))){
    sdf = df # subset(df, !is.na(df$norm))
    # Fit a local polynomial model to the normalized, centered data per plate.
    m = locfit(norm ~ lp(xpos, ypos), data = sdf, lfproc=locfit.robust)
    # Subtract the polynomial from the normalized data
    dd$cor.loc[id_select] = df$norm - predict(m, newdata=df)
    # loess model:
    if("pattern" %in% names(dd)){
      for(jj in seq_along(region)) {
        mjj = loess(norm ~ row + col, data = sdf[sdf$pattern == region[jj], ], span=0.5)
        dd$cor.loess[id_select & dd$pattern == region[jj]] = mjj$y - mjj$fitted 
      }
    } else {
      m = loess(norm ~ row + col, data = sdf, span=0.5)
      dd$cor.loess[id_select] = m$y - m$fitted 
    }
  }
}


```


```{r}
#| label: correction_comparison

tmp = dd[dd$plate %in% plates_included,]
p1 = ggplot(tmp, aes(x=cor, y=cor.loess, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas + plate, ncol=2) + xlab("Median polish")
p2 = ggplot(tmp, aes(x=cor.loc, y=cor.loess, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~ meas + plate, ncol=2)
#plot_grid(p1, p2, nrow=2, labels = "AUTO")
ggarrange(p1, p2, nrow=2, labels = "AUTO", common.legend = T, legend = "right")

if(geom_cor=="median polish") dd$cor = dd$cor.mp
if(geom_cor=="loess") dd$cor = dd$cor.loess
if(geom_cor=="local polynomial fit") dd$cor = dd$cor.loc
```


```{r}
#| label: fig-norm_cor_row
#| fig-show: asis
#| fig-width: 12
#| fig-height: 12
#| fig-cap: "Normalized (A) and geometry-corrected signal (B) versus row number. Systematic row effects should not be visible any more."
#| fig-dpi: 300

tmp = dd[dd$plate %in% plates_included,]
p1 = ggplot(tmp, aes(x=row, y=norm, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas + plate, ncol=2)
p2 = ggplot(tmp, aes(x=col, y=norm, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas + plate, ncol=2)
p3 = ggplot(tmp, aes(x=row, y=cor, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas + plate, ncol=2)
p4 = ggplot(tmp, aes(x=col, y=cor, color=cpd_type)) + geom_point() +
  scale_color_manual(name="", values=type.pal, drop=F) +
  facet_wrap(~meas + plate, ncol=2)
Sys.sleep(5)
ggarrange(p1, p3, nrow=2, ncol=1, labels = "AUTO", 
          common.legend = T, legend = "right")
Sys.sleep(5)
```


```{r}
#| label: fig-norm_cor_col
#| fig-show: asis
#| fig-width: 12
#| fig-height: 12
#| fig-cap: "Normalized (A) and geometry-corrected signal (B) versus column number. Systematic column effects should not be visible any more."
#| fig-dpi: 300

ggarrange(p2, p4, nrow=2, ncol=1, labels = "AUTO", 
          common.legend = T, legend = "right")
Sys.sleep(5)
```


```{r}
#| label: plate_hmap

p1 = ggplate.hm(dd, "raw") + ggtitle("Raw signal")
p2 = ggplate.hm(dd, "norm") + ggtitle("Normalized signal")
p3 = ggplate.hm(dd, "cor") + ggtitle("Geometry corrected signal")

## some useful condition how to order plates
## 1. multiple conditions, multiple celltypes:
if(nr_conditions * nr_cell_types > 1) {
  height = 4 + 2.5 * nr_lib_plates
  width = 4 + 3.5 * nr_conditions*nr_cell_types
  p1 = p1  + facet_grid(dd$lib_plate ~ dd$cell_type + dd$condition)
  p2 = p2  + facet_grid(dd$lib_plate ~ dd$cell_type + dd$condition)
  p3 = p3  + facet_grid(dd$lib_plate ~ dd$cell_type + dd$condition)
}
## 2. single condition, single celltype, many plates:
if(nr_conditions * nr_cell_types == 1) {
  height = 3 + 2.5 * sqrt(nr_lib_plates)
  width = 6 + 4.5 * sqrt(nr_lib_plates)
  nr_columns = floor(1.5 * sqrt(nr_lib_plates))
  p1 = p1  + facet_wrap(~dd$lib_plate, ncol = nr_columns)
  p2 = p2  + facet_wrap(~dd$lib_plate, ncol = nr_columns)
  p3 = p3  + facet_wrap(~dd$lib_plate, ncol = nr_columns)
}


if(!file.exists(path_output%&% "plate_hmap_raw.png")) {
  ggsave(path_output%&% "plate_hmap_raw.png", p1, units = "cm", 
       width=width, height=height, dpi = 300)
} 
if(!file.exists(path_output%&% "plate_hmap_norm.png")) {
  ggsave(path_output%&% "plate_hmap_norm.png", p2, units = "cm", 
       width=width, height=height, dpi = 300)
}
if(!file.exists(path_output%&% "plate_hmap_cor.png")) {
  ggsave(path_output%&% "plate_hmap_cor.png", p3, units = "cm", 
       width=width, height=height, dpi = 300)
}

```

Plate heatmaps of raw, normalized, and corrected signal are stored as image files. 
Based on good practice and visual inspection, we decided to apply `r geom_cor` correction.


## Plate similarity & outlier detection


Multidimensional scaling allows to visualize the (dis-) similarity between plates in two dimensions for convenient plotting (@fig-mds). For less than 3 data points, the plot is empty.


```{r}
#| label: fig-mds
#| fig-show: asis
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Plate similarity in a 2D projection by multi-dimensional scaling. "
#| fig-dpi: 300

tmp = dd[order(dd$plate, dd$row, dd$col),]
mdat = acast(tmp[tmp$cpd_type == "C", ], well ~ plate + meas, value.var = "cor")
mdat[which(!is.finite(mdat))] = 0
sampleDist = dist( t( mdat ) )
#attr(sampleDist, "Labels") = as.character(meta$plate)
sampleDistMatrix = as.matrix( sampleDist )
if(nrow(sampleDistMatrix) < 3){
  d.mds = data.frame(V1=0, V2=0)
  t.range = c(-1,1)
  t.color="none"
  t.label="EMPTY"
  percentVar = c(0,0)
} else {
  d.mds = cmdscale(sampleDist, k=(nrow(sampleDistMatrix)-1), eig = T, add=T)
  percentVar = d.mds$eig/sum(d.mds$eig)
  d.mds = as.data.frame(d.mds$points)
  t.range = 1.5*c(min(d.mds[,1:2]), max(d.mds[,1:2]))
  t.label = gsub("(.*)_" %&% dd$meas[1], "\\1", colnames(mdat))
  #t.color = gsub(".*_(.*)_" %&% dd$meas[1], "\\1", colnames(mdat))
  t.color = meta_plates$condition[match(t.label, meta_plates$plate)]
}
ggplot(d.mds, aes(x=V1, y=V2, col=t.color)) + 
  geom_text(aes(x=V1, y = V2, label=t.label), 
            size=3, show.legend=F) + 
  geom_point(size=-1) +
  scale_colour_brewer(name = "Condition", palette="Dark2") + 
  xlim(t.range) + ylim(t.range) +
  guides(colour = guide_legend(override.aes = list(size=3, shape=15)))+  
  xlab(paste0("PC1: ", round(100* percentVar[1]), "% variance")) +
  ylab(paste0("PC2: ", round(100* percentVar[2]), "% variance")) 

```




# Differential Activity Analysis


```{r}
#| label: z_test1

# replace plate stats with plate stats from "cor" signal
plate_stats = get_plateStats(dd, "cor")
tmp = melt(plate_stats, measure.vars = c("med", "mad"))
plate_stats = dcast(tmp, plate + meas ~ variable + cpd_type, value.var = "value")
plate_stats = left_join(plate_stats, unique(dd[, c("plate", "cell_type")]))
id_keep = !grepl("med|mad", names(dd))
dd = left_join(dd[, id_keep], plate_stats)

# use only negative controls (or cpds)
t.med = dd$med_N
t.mad = dd$mad_N
t.med.id = "negative controls"
t.mad.id = "negative controls"
t.med.diff = abs(median(dd$med_N, na.rm=T) - median(dd$med_C, na.rm=T))
t.mad.diff = abs(median(dd$mad_N, na.rm=T) - median(dd$mad_C, na.rm=T))
if (t.med.diff > 0.02) {
  t.med = dd$med_C
  t.med.id = "compounds"
}
if (t.mad.diff > 0.02) {
  t.mad = dd$mad_C
  t.mad.id = "compounds"
} 

dd$z = (dd$cor - t.med) / t.mad
dd = dd[is.finite(dd$z), ]
dd$pval = 2 - 2 * pnorm(abs(dd$z))

```



```{r}
#| label: neg_dist1
#| fig-dpi: 300

## density plot
neg = dd[dd$cpd_type=="N",]
quant = dnorm(neg$cor, mean=median(neg$cor), sd=mad(neg$cor))
p1 = ggplot(dd, aes(cor, after_stat(density), group=cpd_type, color=cpd_type)) + 
  stat_density(linewidth=1, show.legend=T, geom="line", position="identity") +
  scale_color_manual(name="", values=type.pal, drop=F) +
  scale_x_continuous(limits=c(-1,2)) +
  guides(color = guide_legend(override.aes = list(size=3))) +
  labs(x="Corrected signal", y="Density") + ggtitle("Density Plot")
## qq-plot
# quantile x group matrix:
y = matrix(unlist(tapply(dd$cor, dd$cpd_type, quantile, c(0.25,0.75), na.rm=T)), 
           ncol=2, byrow=T)
x = qnorm( c(0.25, 0.75))         # Find the matching normal values on the x-axis
slope = t(diff(t(y))) / diff(x)   # Compute the line slope
int = y[,1] - slope * x[1]        # Compute the line intercept
# assemble the qqline data frame:
qqline = data.frame(int=int, slope=slope, type=levels(droplevels(dd$cpd_type)))
levels(qqline$type) = levels(dd$cpd_type)
# do the plot
p2 = ggplot(dd, aes(sample=cor, group=cpd_type, color=cpd_type)) + stat_qq(distribution = qnorm) +
  geom_abline(data = qqline, aes(intercept=int, slope=slope)) + 
  scale_color_manual(name="", values=type.pal, drop=F) +
  labs(x="Theoretical Quantiles", y="Signal Quantiles") + ggtitle("Normal QQ Plot")
## MvA plot, 1 point for each compound type on each plate
meas.vars = grep("med|mad", names(plate_stats))
tmp = melt(plate_stats, id.vars = 1, measure.vars = meas.vars)
tmp$cpd_type = gsub(".*_(.*)", "\\1", tmp$variable)
tmp$cpd_type = factor(tmp$cpd_type, levels=levels(dd$cpd_type))
tmp$variable = gsub("(.*)_.*", "\\1", tmp$variable)
tmp = dcast(tmp, plate + cpd_type ~ variable)
p3 = ggplot(tmp, aes(x=med, y=mad, group=cpd_type, color=cpd_type)) + 
  geom_jitter(width=rel(0.05)) + labs(x="Median", y="MAD") + 
  #scale_y_continuous(limits=c(0,1)) +
  #scale_x_continuous(limits=c(-0.1,1.1), breaks=seq(0,1,0.2)) + 
  scale_color_manual(name="", values=type.pal, drop=F) + ggtitle("Mean Variance Plot")
```



The distributions of normalized and corrected signals for all controls and compounds in @fig-neg_dist is a useful illustration for the correspondence between model assumptions and the data. Based on a careful examination of the distributions, the plate-wise median of the `r t.med.id` and the plate-wise MAD of the `r t.mad.id` was used for testing.


```{r}
#| label: fig-neg_dist
#| fig-show: asis
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Distribution of controls and compounds. (A) Empirical probability density, (B) Quantile-Quantile plot, (C) plate-wise MAD vs plate-wise median. (D) P-value distribution. Regular PVD's can have 2 distinct shapes: (i) uniform if there is no DEG (plateau at density = 1). (ii) a peak at small p-values and a plateau < 1 for p -> 1; the plateau value corresponds to the proportion of DEG's. Any other shape is irregular and sensitively indicates a violation of the model assumptions."
#| fig-dpi: 300

p4 = ggplot(dd[dd$cpd_type=="C", ], aes(pval, after_stat(density))) + 
  geom_histogram(bins=100, fill="grey50", col="white") + 
  geom_hline(yintercept = 1, col="black") + xlab("P-value") + ylab("Density") + 
  theme(legend.position="none") + #coord_cartesian(ylim=c(0, 5)) + 
  ggtitle("PVD Plot")

#plot_grid(p1,p2,p3,p4, nrow=2, labels = "AUTO")MM
ggarrange(p1, p2, p3, p4, nrow=2, ncol=2, labels = "AUTO", 
          common.legend = T, legend = "right")
```





The volcano-plots in @fig-volcano11 display the relation between the significance of the test, expressed as the negative logarithm of the false discovery rate (FDR), and the effect size, expressed as the logarithm of the fold-change. 

```{r}
volcano_caption = "Volcano plots. The horizontal line corresponds to FDR = " %&% fdr_cut %&% " and genes below this line are labeled 'not significant'. The two vertical lines correspond to a " %&% act_cut %&% "-fold change in expression and genes outside this range are labeled 'strong'. The legend refers to 'not' - not significant and weak effect; 'signif & strong' - significant and strong effect; 'signif' - significant but weak effect; 'strong' - not significant but strong effect."

```


```{r}
#| label: fig-volcano11
#| fig-show: asis
#| fig-width: 10
#| fig-height: 5
#| fig-cap: !expr volcano_caption
#| fig-dpi: 300

dd$padj = p.adjust(dd$pval, method="BH")
dd$padj[dd$padj==0] = min(dd$padj[dd$padj>0])
dd$hit = c("not","signif")[as.numeric(dd$padj < fdr_cut)+1]
dd$hit[abs(dd$cor) > (act_cut) & dd$hit== "signif"] = "signif & strong"
dd$hit[abs(dd$cor) > (act_cut) & dd$hit== "not"] = "strong"
dd$hit = factor(dd$hit, levels=c("not", "strong", "signif", "signif & strong"))

ggplot(dd[!is.na(dd$hit),], aes(x=cor, y=-log10(padj), col=cpd_type)) + 
  geom_point(shape=16, size=0.9) + 
  geom_hline(yintercept = -log10(fdr_cut)) + 
  geom_vline(xintercept = c(-1,1)*(act_cut)) +
  xlab("Effect size [normalized]") + 
  ylab(expression(paste("Significance [", -log[10](padj), "]"))) +
  scale_color_manual(values = type.pal, name="Cpd type", drop=F) +
  #theme(legend.justification=c(0,1), legend.position=c(0.7,0.35)) +
  theme(legend.justification=c(0,1), legend.position="right") +
  guides(color = guide_legend(override.aes = list(size=3, shape=15))) +
  facet_wrap(~condition, scales="free")

```



```{r}
#| label: tbl-deasum_ctr1
#| results: asis
#| eval: true
#| tbl-cap: "DAA summary for the control and reference compounds."

dd$direction = as.character(dd$hit)
dd$direction[dd$cor > 0] = dd$hit[dd$cor > 0] %&% " up"
dd$direction[dd$cor < 0] = dd$hit[dd$cor < 0] %&% " down"
dd$direction = factor(dd$direction, 
                        levels=c("signif & strong down", "signif down",
                                 "strong down", "not down",
                                 "not up", "strong up",
                                 "signif up", "signif & strong up"))
tmp = dd[dd$cpd_type != "C",]
tmp$cpd_type = droplevels(tmp$cpd_type)
m = table(tmp$direction, tmp$cpd_type)
#kable(addmargins(m, margin = 1))
kable(t(as.matrix(m)))
```


```{r}
#| label: fig-volcano21
#| fig-show: asis
#| fig-width: 10
#| fig-height: 5
#| fig-cap: !expr volcano_caption
#| fig-dpi: 300

res = dd[dd$cpd_type=="C",]
#res$lib_plate = droplevels(res$lib_plate)
res$padj = p.adjust(res$pval, method="BH")
res$padj[res$padj==0] = min(res$padj[res$padj>0])
res$hit = c("not","signif")[as.numeric(res$padj < fdr_cut)+1]
res$hit[abs(res$cor) > act_cut & res$hit== "signif"] = "signif & strong"
res$hit[abs(res$cor) > act_cut & res$hit== "not"] = "strong"
res$hit = factor(res$hit, levels=c("not", "strong", "signif", "signif & strong"))
res$direction = as.character(res$hit)
res$direction[res$cor > 0] = res$hit[res$cor > 0] %&% " up"
res$direction[res$cor < 0] = res$hit[res$cor < 0] %&% " down"
res$direction = factor(res$direction, 
                        levels=c("signif & strong down", "signif down",
                                 "strong down", "not down",
                                 "not up", "strong up",
                                 "signif up", "signif & strong up"))

cbPalette = c("#000000", "#009E73", "#56B4E9", "#D55E00", 
              "#F0E442", "#0072B2", "#E69F00", "#CC79A7")
res$uwid = res$lib_plate %&% "-" %&% res$well
hit_label = res$name
hit_label[res$hit != "signif & strong"] = NA
ggplot(res, aes(x=cor, y=-log10(padj), col=hit, label=hit_label)) + 
  geom_hline(yintercept = -log10(fdr_cut)) + 
  geom_vline(xintercept = c(-1,1)*(act_cut)) +
  geom_point(shape=16, size=0.9, show.legend = T) + 
  #geom_label_repel(show.legend = F) +
  xlab("Effect size [normalized]") + 
  ylab(expression(paste("Significance [", -log[10](padj), "]"))) +
  scale_color_manual(values = cbPalette, name="DAA class", drop=F) +
  #scale_x_continuous(limits = c(-1,1.5)) +
  scale_y_continuous(limits = c(-0.1,12)) +
  #theme(legend.justification=c(0,1), legend.position=c(0.7,0.35)) +
  theme(legend.justification=c(0,1), legend.position="right") +
  guides(color = guide_legend(override.aes = list(size=3)))+
  facet_wrap(~condition, scales="free")
```

```{r}
#| label: tbl-deasum_cpd1
#| results: asis
#| tbl-cap: "DAA summary for the test compounds."


m = table(res$condition, res$direction)
#kable(addmargins(m, margin = 1))
kable(as.matrix(m))

```

In summary, among the `r nrow(res)` tested compounds, `r length(which(as.numeric(res$hit)>2))` were found to be differentially active and `r length(which(res$hit=="signif & strong"))` additionally showed strong activity above the chosen threshold. This corresponds to an overall hit rate of `r signif(length(which(as.numeric(res$hit)>2))/nrow(res)*100,2)` percent. An overview on the test results is given in @tab-deasum_cpd1.



# Description of result files

Results are stored in tab-delimited text file format: one file containing the normalized, corrected readout of all wells including the controls (`DAA_allwells.tsv`) and one file containing the DAA results of the significantly and strongly active compounds with substance name and physico-chemical properties and descriptors added (`DAA_results.tsv`). The description of the exported columns is described in @tab-daares and in the documentation of the library supplier.



```{r}
#| label: tbl-daares
#| results: asis
#| tbl-cap: "Description of `DAA_results.txt` file header."

col.id = c("lib_plate", "plate", "well", 
           "raw", "norm", "cor", 
           "z", "pval", "padj", 
           "direction", "compound")
m = matrix( c(col.id, "Library plate", "Assay plate barcode", 
              "Well ID", "raw signal", "normalized signal", 
              "corrected signal", "Z-statistics", "two-sided raw p-value", 
              "two-sided adjusted p-value (=FDR)", "direction of the effect",
              "Compound name"), 
            ncol=2, byrow=F)
colnames(m) = c("Name","Description")
kable(m)
```



```{r}
#| label: save11
#| eval: true

# save plate statistics:
plate_stats$z_prime = 1 - 3*(plate_stats$mad_N + plate_stats$mad_P) / 
  abs(plate_stats$med_P - plate_stats$med_N)
plate_stats$SSMD = abs(plate_stats$med_P - plate_stats$med_N) / 
  sqrt(plate_stats$mad_N^2 + plate_stats$mad_P^2)
write.table(plate_stats, file = path_output %&% "plate_stats.tsv", 
            sep="\t", row.names = F, na = "")

# save all wells w/o FDR or cpd info
out = dd[order(dd$cell_type, dd$lib_plate, dd$plate, dd$well, decreasing=F), ]
out = out[!is.na(dd$padj), ]
write.table(out[, col.id[1:9]], file = path_output %&% "DAA_allwells.tsv", 
            sep="\t", row.names = F, na = "")

# save only hits w/ FDR & cpd info
out = res[which(res$hit == "signif & strong"), c("uid", col.id[1:10])]
lib_desc = read.csv(path_libfile)
lib_desc = lib_desc[, c(1, 10:ncol(lib_desc))]
out = left_join(out, lib_desc, by = c("uid" = "unique_identifier"))
write.table(out[, -1], file = path_output %&% "DAA_results.tsv", 
            sep="\t", row.names = F, na = "")
```


